Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-18 BN_discretizer_paper.tex
woo

\begin{table}[H]
  \centering
  \begin{tabular}{@{}ccl@{}}
    \toprule
    Current & Mykel & Description \\
    \midrule
    % $G$ & $G$ & BN graph structure \\
    % $P$ & $P$ & discrete probability \\
    % $p$ & $p$ & probability density \\
    % $D$ & $D$ & available data \\
    $|D|$ & $n$ & number of variables in the BN \\
    $D_{\bm{Y}}$ & & subset of data associated with $\bm{Y}$ \\
    $X$ & $X_{1:n}$ & variables in BN (discrete) \\
    $x_{1:n}$ & $x_{1:n}$ & a specific instance of vars in BN \\
    $x'_{1:n}$ &  & unique instances of var in BN \\
        & $r_i$ & number of instantiations of $i$th var \\
    $J_p$ & $q_i$ & number of instantiations of parents of $i$th var \\
        &       & $ = 1$ if no parents \\
    $\brock{\Pi_{X_i}}_j$ & $\pi_{ij}$ & $j$th instantiation of $i$th var \\
        & $\theta_{ijk}$ & probability of $i$th var $= k$ given $j$th parental inst. \\
    $\Theta$ & $\bm{\theta}$ & set of all (non-independent) probability parameters \\
    $n_{i,j}^{(k)}$ & $m_{ijk}$ & observation counts \\
    $n_{i,j,l}^{(k)}$ & $m_{ijk}$ & observation counts with $l$th instance of spouse\\
    $\alpha_{ijk}$ & $\alpha_{ijk}$ & Dirichlet prior counts \\
    $N_{ij}$ & $\alpha_{ij0}$ & sum over Dirichlet prior counts for $k$ \\
        & $m_{ij0}$ & sum over observations counts for $k$ \\
        & $\Gamma$ & the Gamma function \\
    $N$ &          & is the number of samples in dataset \\
    $N_i$ &          & counts for $i$th instance of a specific var \\
    $\Pi_X$ & $\text{pa}_{x_i}$ & parents of a variable or instance \\
    $n_C$ & & number of children of a variable \\
    $H(~)$ & & a function \\
    $I(~,~)$ & & mutual information \\
    $X_i$ & & $i$th variable in BN, may be cont or disc \\
    $X_i^*$ & & discrete version of cont var \\
    $E$ & & set of edges in BN \\
    $\text{Pa}\paren{X_i}$, $n_p$ & & number of parents of $X_i$ \\
    $f\paren{X_i, \Pi_{X_i}}$ & $f(G)$ & score function for K2 \\
    $M_C$ & & discretization policy for $C$ \\
    $C$ & & a continuous variable \\
    $t_i$ & & the $i$th bin edge for a specific var \\
    $d_i$ & & allowable discretization positions (between unique vals) \\
    $k$ & & number of intervals for $M_C$ \\
    $C_{1:N}$ & & sorted samples for $C$ \\
    $n_i$ & & $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$ \\
    $s_i$ & & $\sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$ \\
    $C_i$ & & $i$th child of a specific var \\
    $\boldsymbol{S_i}$ & & set of spouses associated with $i$th child of specific var \\
    $L$ & & largest cardinality among discrete variables in BN \\
    $I_q$ & & an interval from $u \rightarrow v$ \\
    $v'_\text{max}$ & & largest cardinality of variables that directly connect to $X$ \\
    $S[]$ & &  optimal objective value of subproblems with instances from $1$ to $u$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

There are three common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters.
One successful example are Gaussian graphical models \citep{Weiss_2011}.
For other parametric distributions and non-linear functions \citep{Ihler_2009}, these redesigned algorithms are computationally expensive and do not perform well.
The second is to use nonparametric distributions, such as particle representations and Gaussian processes \todo{[cite]}.
\todo{These have the advantage of BLAH BLAH, and their parameter set grows lineary with the amount of data.}

@@

The MDL principle was first proposed by \cite{MDL_1978} and states that the best model for a dataset is one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description lengths of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.
\todo{Results in this work suggest that the MDL method suffers from....}

@@

\todo{
Introduce P, C, and S for parents
}

@@

It is often necessary to infer the structure of a Bayesian network from data.
Three common approaches to Bayesian network structure learning are constraint-based, score-based, and Bayesian model averaging \citep[see][chap.~18]{PGM_2009}.
This work uses the K2 structure learning algorithm \citep{K2}, one of the most successful score-based structure learning methods.
Score-based structure learning methods over discrete variables commonly evaluate candidate structures according to their likelihood against a dataset $D$.
Let $m_{ij}^{(k)}$ be the number of times $X_i = k$ given $\bm{\pi}_{ij}$ in $D$.
The likelihood is given by

\begin{equation}
P(D\mid \bm{\theta}, G) = \prod_{i=1}^n \prod_{j=1}^{q_i} \prod_{k=1}^{r_i} \brock{\theta_{ij}^{(k)}}^{m_{ij}^{(k)}}\text{.}
\end{equation}

@@

\begin{small}
\begin{equation}
\label{eq:BayesianScore}
  \ln P(D\mid G) = \ln P(G) + \sum_{i=1}^n \sum_{j=1}^{q_i} \ln\paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(0)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(0)} + m_{ij}^{(0)}}
    }
  } + \sum_{k=1}^{r_i} \ln\paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(k)} + m_{ij}^{(k)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(k)}}
    }
  }\text{.}
\end{equation}
\end{small}

@@

\begin{equation}
\alpha_{ij}^{(0)} = \sum_{k=1}^{r_i} \alpha_{ij}^{(k)} \quad \qquad m_{ij}^{(0)} = \sum_{k=1}^{r_i} m_{ij}^{(k)}\text{.}
\end{equation}

@@

\noindent
The discretization policy discretizes $\cX$ into $k$ intervals.
Let the samples of $\cX$ in a given dataset $D$ be sorted in ascending order, ${\cx_{1:n} = \curly{\cx_1 \leq \cx_2 \leq \ldots \leq \cx_n}}$, and let the unique values be ${\cu_{1:n} = \curly{\cu_1 \leq \cu_2 \leq \ldots \leq \cu_n}}$.
The index of the last occurence of $\cu_i$ in $\cx$ is denoted $s_i$.

The discretization edges $e_{1:k-1}$ mark the boundaries between discretization intervals.
In this work they are restricted to the midpoints between unique ascending instances of $\cX$.
Thus, each edge $e_i = \paren{\cu_{s_{j+1}} - \cu_{s_j}}/2$ for some $s_j$.
An integer representation of $\Lambda_{\cX}$ can thus be written $\Lambda_{\cX} = \curly{\lambda_1 < \lambda_2  < \ldots < \lambda_k}$, where $\lambda_1 = s_1$ and $\lambda_i \in s_{1:|\cu|}$.

@@

This section covers the discretization of a single continuous variable $\cX$ in a Bayesian network where all other variables are discrete.
Let $n_p$ and $n_c$ respectively refer to the number of parents and number of children of $X$.
Each child $C_i$ of $X$ may possess other parents, referred to as spouses of $X$.
The set of spouses of $X$ is denoted $\boldsymbol{S_i}$.

Let $D$ be a dataset of $N$ samples from which the discretization will be learned, sorted in ascending order according to $X$.
Let $D_{\boldsymbol{Y}}$ will refer to the data instances associated with variables $\boldsymbol{Y}$, thus $D_X = \{ x_1,x_2,x_3,\ldots,x_N \}$.
In order to facilitate understanding, we temporarily assume all values in $D_X$ are unique, that is, $x_1 < x_2 < \ldots < x_N$.
It follows that a discretization policy $M$ on $X$ can be written as $M_X = \brock{n_1,n_2,\ldots,n_k}$, where $k$, $n_1$, $n_2$, $\ldots,n_k$ are positive integers satisfying $N = \sum_{i=1}^k n_i$.
\mykel{$\bigstar$ This conflicts with Section 3.2, right?}
An optimal discretization policy maximizes $P(M)\cdot P(D\mid M)$ for some prior $P(M)$ and likelihood $P(D\mid M)$.

@@

The following four principles for the optimal discretization policy, motivated by MODL, enable the formulation of $P(M)$ and $P(D \mid M)$:
\begin{enumerate}
\item A discretization edge located at $(x_i + x_{i+1})/2$ has probability
\begin{equation}
1 - {\exp(- L \cdot {{x_{i+1} - x_i}\over{x_N - x_1}})},
\end{equation}
where $L$ is the largest cardinality number of discrete variables in $X$'s Markov blanket. \tim{Where does this come from? Is there an intuitive explanation?}\mykel{$\bigstar$}
\item For a given interval of $M$, every distribution of $\Pi_X$ is equiprobable.
\item For each pair of $(C_i,\boldsymbol{S_i})$ and a given interval of $M$, every distribution of $C_i$ with a value of $\boldsymbol{S_i}$ is equiprobable.\mykel{$\bigstar$}
\item The distributions of each $C_i$ and $\Pi_X$ in each interval of $M$ with each value of $\boldsymbol{S_i}$ are independent from each other.\mykel{$\bigstar$}
\end{enumerate}

From the first principle one obtains
\begin{equation}
  \label{eq:p_M}
  P(M) = \prod_{i=1}^k \curly{ \brock{1 - {\exp(- L \cdot {{x_{s_i+1} - x_{s_i}}\over{x_N - x_1}})}} \cdot \exp\paren{L \cdot {{x_{s_{i}} - x_{s_{i-1} + 1}}\over{x_N - x_1}}}},
\end{equation}

\noindent
where $s_i = \sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$.
The Bayesian network graph structure causes the likelihood term $P(D \mid M)$ to factorizes according to $X$'s Markov blanket:

\begin{equation}
  \label{eq:p_D_given_M}
  P\paren{D \mid M} \propto P\paren{D_{\Pi_x} \mid M} \cdot \prod_{i = 1}^{n_c} P\paren{D_{C_i} \mid M, D_{\boldsymbol{S_i}}}\text{.}
\end{equation}

For example, in Figure~\ref{fig:example_factorization}, the corresponding $P(D \mid M)$ is factorized according to
\begin{equation}
P\paren{D_{ P_1,P_2,P_3 } \mid M} \cdot P\paren{ D_{ C_1 } \mid M,D_{\boldsymbol{S_1}}} \cdot P\paren{D_{C_2} \mid M,D_{ S_2  }}\text{.}
\end{equation}

\begin{figure}[ht]
  \label{fig:example_factorization}
  \begin{tabular}{cc}
    \input{graph1}
  \end{tabular}
  \caption{Factorization of $P(D \mid M)$}
\end{figure}

The concept behind the factorization also the motivation behind forward sampling in a Bayesian network.
The parents of $X$ are independent of the children given $X$.
The parents are not necessarily individually independent, and thus the parental term $P(D_{\Pi_x} \mid M)$ cannot be factored further.
The children of $X$ are similarly independent given $X$ and its spouses $\boldsymbol{S_i}$, leading to their factored product $\prod_{i = 1}^{n_c} P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$.
Each component in the decomposition be evaluated based on a discretization policy $M = [n_1,n_2,\ldots,n_k]$ and dataset $D$.

\subsubsection{Evaluation of $P(D_{\Pi_x} \mid M)$}
Let $J_P = \prod_{i=1}^{n_p} \| P_i \|$. It follows that

\begin{equation}
\label{eq:likelihood_one}
P(D_{\Pi_X} \mid M) = \prod_{i=1}^k  {{1}\over{{n_i + J_P - 1}\choose{J_P -1}}}
{{1}\over{ {{{n}_i}!}\over{ {n^{(p)}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_P} !}}  }}\text{,}
\end{equation}

@@

\subsubsection{Evaluation of $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$}
For each pair $(C_j, \boldsymbol{S_j})$, let $\|C_j \| = J_j$ and $\| \boldsymbol{S_j} \| = L_j = \prod_{v \in \boldsymbol{S_j}} \| v \|$. It follows that

\begin{equation}
\label{eq:likelihood_two}
P(D_{C_j}  \mid M, D_{S_j}) =
\prod_{i=1}^{k} \prod_{l=1}^{L_j} {{1}\over{{n_{i,l} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}}  }}\text{,}
\end{equation}

@@

\begin{equation}
\label{eq:likelihood_three}
P(D_{C_j}  \mid M) =
\prod_{i=1}^{k}  {{1}\over{{n_{i} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i}}!}\over{ {n^{(j)}_{i,1,\emptyset} !} {n^{(j)}_{i,2,\emptyset} !} \cdots {n^{(j)}_{i,J_j,\emptyset} !}}  }}\text{,}
\end{equation}

@@

The objective function can be formulated given equations \ref{eq:p_M}, \ref{eq:p_D_given_M}, \ref{eq:likelihood_one}, and \ref{eq:likelihood_three}.
The log-inverse of $P(M) \cdot P(D|M)$ is minimized for computational convenience:

@@

Every midpoint between two nonequal consecutive values in $D_X$ is a candidate discretization edge.
Let ${D'_X} = \brock{x'_1,x'_2,\ldots,x'_M}$ be the set of unique values of $D_X$ sorted in ascending order, where $M$ is the number unique values in $D_X$.
Recall that $D_X$ is also sorted.
Let $b = \brock{b_0,b_1,b_2,\ldots,b_M}$ be an increasing sequence of integers marking the repeated value intervals in $D_X$ such that $b_0 = 0$ and $x_{b_{i-1} + 1} = x_{b_{i-1} + 2} = \ldots = x_{b_i} = x'_i$.
By this definition, the allowable discretization positions are $d_i = \paren{x_{b_{i}} + x_{{b_i}+1}}/2$ for all \todo{$i = 1,2,\ldots,M-1$}.

@@

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm~\ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $M$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\State
\For {$v = 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u = 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $M$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\section{Experiments}
\label{sec:experiments}

This section describes experiments conducted to evaluate the proposed methods.
All experiments were run on datasets from the publically available Irvine machine learning repository \citep{Lichman_2013}.
Variables are labelled alphabetically in the order given on the dataset information webpage.
Subscripts inducate the number of discretized intervals.
Shaded nodes are originally discrete and unshaded nodes are originally continuous.

>>>>

% \begin{table}[H]
%   \centering
%   \begin{tabular}{@{}ccl@{}}
%     \toprule
%     Current & Mykel & Description \\
%     \midrule
%     % $G$ & $G$ & BN graph structure \\
%     % $P$ & $P$ & discrete probability \\
%     % $p$ & $p$ & probability density \\
%     % $D$ & $D$ & available data \\
%     $|D|$ & $n$ & number of variables in the BN \\
%     $D_{\bm{Y}}$ & & subset of data associated with $\bm{Y}$ \\
%     $X$ & $X_{1:n}$ & variables in BN (discrete) \\
%     $x_{1:n}$ & $x_{1:n}$ & a specific instance of vars in BN \\
%     $x'_{1:n}$ &  & unique instances of var in BN \\
%         & $r_i$ & number of instantiations of $i$th var \\
%     $J_p$ & $q_i$ & number of instantiations of parents of $i$th var \\
%         &       & $ = 1$ if no parents \\
%     $\brock{\Pi_{X_i}}_j$ & $\pi_{ij}$ & $j$th instantiation of $i$th var \\
%         & $\theta_{ijk}$ & probability of $i$th var $= k$ given $j$th parental inst. \\
%     $\Theta$ & $\bm{\theta}$ & set of all (non-independent) probability parameters \\
%     $n_{i,j}^{(k)}$ & $m_{ijk}$ & observation counts \\
%     $n_{i,j,l}^{(k)}$ & $m_{ijk}$ & observation counts with $l$th instance of spouse\\
%     $\alpha_{ijk}$ & $\alpha_{ijk}$ & Dirichlet prior counts \\
%     $N_{ij}$ & $\alpha_{ij0}$ & sum over Dirichlet prior counts for $k$ \\
%         & $m_{ij0}$ & sum over observations counts for $k$ \\
%         & $\Gamma$ & the Gamma function \\
%     $N$ &          & is the number of samples in dataset \\
%     $N_i$ &          & counts for $i$th instance of a specific var \\
%     $\Pi_X$ & $\text{pa}_{x_i}$ & parents of a variable or instance \\
%     $n_C$ & & number of children of a variable \\
%     $H(~)$ & & a function \\
%     $I(~,~)$ & & mutual information \\
%     $X_i$ & & $i$th variable in BN, may be cont or disc \\
%     $X_i^*$ & & discrete version of cont var \\
%     $E$ & & set of edges in BN \\
%     $\text{Pa}\paren{X_i}$, $n_p$ & & number of parents of $X_i$ \\
%     $f\paren{X_i, \Pi_{X_i}}$ & $f(G)$ & score function for K2 \\
%     $M_C$ & & discretization policy for $C$ \\
%     $C$ & & a continuous variable \\
%     $t_i$ & & the $i$th bin edge for a specific var \\
%     $d_i$ & & allowable discretization positions (between unique vals) \\
%     $k$ & & number of intervals for $M_C$ \\
%     $C_{1:N}$ & & sorted samples for $C$ \\
%     $n_i$ & & $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$ \\
%     $s_i$ & & $\sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$ \\
%     $C_i$ & & $i$th child of a specific var \\
%     $\bm{S}_i$ & & set of spouses associated with $i$th child of specific var \\
%     $L$ & & largest cardinality among discrete variables in BN \\
%     $I_q$ & & an interval from $u \rightarrow v$ \\
%     $v'_\text{max}$ & & largest cardinality of variables that directly connect to $X$ \\
%     $S[]$ & &  optimal objective value of subproblems with instances from $1$ to $u$ \\
%     \bottomrule
%   \end{tabular}
% \end{table}

@@

There are three common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters.
One successful example are Gaussian graphical models \citep{Weiss_2011}.
For other parametric distributions and non-linear functions \citep{Ihler_2009}, these redesigned algorithms are computationally expensive and do not perform well.
The second is to use nonparametric distributions, such as particle representations and Gaussian processes \todo{[cite]}.
Unlike parametric methods, nonparametric methods do not make assumptions about the probability distributions being used.
Parametric models have a fixed number of parameters, whereas the number of parameters in a nonparametric model grow with the amount of training data.

@@

The MDL principle was first proposed by \cite{MDL_1978} and states that the best model for a dataset is one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description lengths of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.
Results in this work suggest that the MDL method suffers from \todo{fill this out.}

@@

@@

It is often necessary to infer the structure of a Bayesian network from data.
Three common approaches to Bayesian network structure learning are constraint-based, score-based, and Bayesian model averaging \citep[see][chap.~18]{PGM_2009}.
This work uses the K2 structure learning algorithm \citep{K2}, one of the most successful score-based structure learning methods.
Score-based structure learning methods over discrete variables commonly evaluate candidate structures according to their likelihood against a dataset $D$.
Let $\beta_{ij}^{(k)}$ be the number of times $X_i = k$ given $\bm{\pi}_{ij}$ in $D$.
The likelihood is given by

\begin{equation}
P(D\mid \bm{\theta}, G) = \prod_{i=1}^n \prod_{j=1}^{q_i} \prod_{k=1}^{r_i} \brock{\theta_{ij}^{(k)}}^{\beta_{ij}^{(k)}}\text{.}
\end{equation}

@@

\begin{small}
\begin{equation}
\label{eq:BayesianScore}
  \ln P(D\mid G) = \ln P(G) + \sum_{i=1}^n \sum_{j=1}^{q_i} \ln\paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(0)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(0)} + \beta_{ij}^{(0)}}
    }
  } + \sum_{k=1}^{r_i} \ln\paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(k)} + \beta_{ij}^{(k)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(k)}}
    }
  }\text{.}
\end{equation}
\end{small}

@@

\begin{equation}
\alpha_{ij}^{(0)} = \sum_{k=1}^{r_i} \alpha_{ij}^{(k)} \quad \qquad \beta_{ij}^{(0)} = \sum_{k=1}^{r_i} \beta_{ij}^{(k)}\text{.}
\end{equation}

@@

\noindent
The discretization policy discretizes $\cX$ into $k$ intervals.
Let the samples of $\cX$ in a given dataset $D$ be sorted in ascending order, ${\cx_{1:n} = \curly{\cx_1 \leq \cx_2 \leq \ldots \leq \cx_n}}$, and let the unique values be ${\cu_{1:m} = \curly{\cu_1 \leq \cu_2 \leq \ldots \leq \cu_m}}$.
The index of the last occurence of $\cu_i$ in $\cx_{1:n}$ is denoted $s_i$.

The discretization edges $e_{1:k-1}$ mark the boundaries between discretization intervals.
In this work they are restricted to the midpoints between unique ascending instances of $\cX$.
Thus, each edge $e_i = \paren{\cu_{s_j} + \cu_{s_{j+1}}}/2$ for some $s_j$.
An integer representation of $\Lambda_{\cX}$ can thus be written $\Lambda_{\cX} = \curly{\lambda_1 < \lambda_2  < \ldots < \lambda_k}$, where $\lambda_1 = s_1$ and $\lambda_i \in s_{1:m}$.

Finally, when discussing the discretization of a particular continuous variable $\cX$, let $P_i$ be the $i$th parent of $\cX$, let $C_i$ be the $i$th child of $\cX$, and let $\bm{S}_i$ be set of spouses of $\cX$ corresponding to the $i$th child.
Context should make it clear when $P$ refers to a discrete probability and when it refers to a variable's parent.

@@

This section covers the discretization of a single continuous variable $\cX$ in a Bayesian network where all other variables are discrete.
An optimal discretization policy for a dataset $D$ maximizes $P(\Lambda)\cdot P(D\mid \Lambda)$ for some prior $P(\Lambda)$ and likelihood $P(D\mid \Lambda)$.

% Let $n_p$ and $n_c$ respectively refer to the number of parents and number of children of $X$.

% Let $D$ be a dataset of $N$ samples from which the discretization will be learned, sorted in ascending order according to $X$.
% Let $D_{\boldsymbol{Y}}$ will refer to the data instances associated with variables $\boldsymbol{Y}$, thus $D_X = \{ x_1,x_2,x_3,\ldots,x_N \}$.
% In order to facilitate understanding, we temporarily assume all values in $D_X$ are unique, that is, $x_1 < x_2 < \ldots < x_N$.
% It follows that a discretization policy $\Lambda$ on $X$ can be written as $M_X = \brock{n_1,n_2,\ldots,n_k}$, where $k$, $n_1$, $n_2$, $\ldots,n_k$ are positive integers satisfying $N = \sum_{i=1}^k n_i$.

@@

Let $D_{\bm{Y}}$ be the subset of the data corresponding to $\bm{Y}$.
Four principles for the optimal discretization policy enable the formulation of $P(\Lambda)$ and $P(D_{-\cX} \mid \Lambda)$, where the probability of the data associated with the target variable $P(D_{\cX})$ is already captured in $P(\Lambda)$.
The four principles, motivated by MODL \todo{and Lustgarden}, are:

\begin{enumerate}
\item The prior probability of a discretization edge between two consecutive unique values $\cu_i$ and $\cu_{i+1}$ is proportional to their difference:

  \begin{equation}
  1 - {\exp(- L \cdot {\frac{\cu_{i+1} - \cu_i}{\cu_m - \cu_1}})}\text{,}
  \end{equation}

where $L$ is the largest number of intervals among discrete variables in $\cX$'s Markov blanket.
This encourages edges between well-separated values over edges between closely packed samples.

\item For a given discretization interval, every distribution over $\Pa_{\cX}$ is equiprobable.
\item For each pair $\anglebrackets{C_i,\bm{S}_i}$ and a given discretization interval, every distribution over $C_i$ given an instance of $\bm{S}_i$ is equiprobable.
\item The distributions over each child and each parent in each discretization interval with each instance of $\bm{S}_i$ are independent from each other.
\end{enumerate}

From the first principle one obtains
\begin{equation}
  \label{eq:p_M}
  P(\Lambda) = \prod_{i=1}^k
  \curly{
    \brock{
      1 - \exp\paren{
        - L \cdot \frac{
                         \cu_{i+1} - \cu_i
                       }{
                         \cu_m - \cu_1
                        }
      }
    } \cdot \exp\paren{
      L \cdot \frac{\cx_{s_{i}} - \cx_{s_{i-1} + 1}}{\cx_n - \cx_1}}
    }\text{.}
\end{equation}

The Bayesian network graph structure causes the likelihood term $P(D \mid \Lambda)$ to factorizes according to $\cX$'s Markov blanket:

\begin{equation}
  \label{eq:p_D_given_M}
  P\paren{D_{-\cX} \mid \Lambda} \propto P\paren{D_{\Pa_\cX} \mid \Lambda} \cdot \prod_{C_i} P\paren{D_{C_i} \mid \Lambda, D_{\bm{S}_i}}\text{.}
\end{equation}

For example, in Figure~\ref{fig:example_factorization}, $P(D_{-\cX} \mid \Lambda)$ factors according to

\begin{equation}
  P\paren{D_{P_1,P_2,P_3} \mid \Lambda} \cdot P\paren{ D_{ C_1 } \mid \Lambda, D_{\bm{S}_1} } \cdot P\paren{D_{C_2} \mid \Lambda, D_{\bm{S}_2}}\text{.}
\end{equation}

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph1}
  \end{tabular}
  \caption{Factorization of $P(D \mid \Lambda)$}
  \label{fig:example_factorization}
\end{figure}

The concept behind the factorization is also the motivation behind forward sampling in a Bayesian network.
The parents of $\cX$ are independent of the children given $\cX$.
The parents are not necessarily individually independent, and thus the parental term $P(D_{\Pa_{\cX}} \mid \Lambda)$ cannot be factored further.
The children of $\cX$ are similarly independent given $\cX$ and the corresponding spouses, leading to their factored product $\prod_{C_i} P(D_{C_i} \mid \Lambda, D_{\bm{S}_i})$.
Each component in the decomposition can be evaluated given a discretization policy and a dataset.

\subsubsection{Evaluation of $P\paren{D_{\Pa_{\cX}} \mid \Lambda}$}

Let $J_P = \prod_{i=1}^{n_p} \| P_i \|$. It follows that

\begin{equation}
  \label{eq:likelihood_one}
  P\paren{D_{\Pa_{\cX}} \mid \Lambda} = \prod_{i=1}^k
    \frac{1}{{{n_i + J_P - 1}\choose{J_P - 1}}}
    \frac{1}{
      \frac{
        {n_i}!
      }{
        {n^{(p)}_{i,1}}! \; {n^{(p)}_{i,2}}! \; \cdots \; {n^{(p)}_{i,J_P}}!
      }
    }\text{,}
\end{equation}

@@

\subsubsection{Evaluation of $P(D_{C_i} \mid \Lambda, D_{\bm{S}_i})$}
For each pair $(C_j, \boldsymbol{S_j})$, let $\|C_j \| = J_j$ and $\| \boldsymbol{S_j} \| = L_j = \prod_{v \in \boldsymbol{S_j}} \| v \|$. It follows that

\begin{equation}
\label{eq:likelihood_two}
P(D_{C_j}  \mid \Lambda, D_{S_j}) =
\prod_{i=1}^{k} \prod_{l=1}^{L_j} {{1}\over{{n_{i,l} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}}  }}\text{,}
\end{equation}

@@

\begin{equation}
\label{eq:likelihood_three}
P(D_{C_j}  \mid \Lambda) =
\prod_{i=1}^{k}  {{1}\over{{n_{i} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i}}!}\over{ {n^{(j)}_{i,1,\emptyset} !} {n^{(j)}_{i,2,\emptyset} !} \cdots {n^{(j)}_{i,J_j,\emptyset} !}}  }}\text{,}
\end{equation}

@@

The objective function can be formulated given equations \ref{eq:p_M}, \ref{eq:p_D_given_M}, \ref{eq:likelihood_one}, and \ref{eq:likelihood_three}.
The log-inverse of $P(\Lambda) \cdot P(D|\Lambda)$ is minimized for computational convenience:

@@

Every midpoint between two nonequal consecutive values in $D_X$ is a candidate discretization edge.
Let ${D'_X} = \brock{x'_1,x'_2,\ldots,x'_\Lambda}$ be the set of unique values of $D_X$ sorted in ascending order, where $\Lambda$ is the number unique values in $D_X$.
Recall that $D_X$ is also sorted.
Let $b = \brock{b_0,b_1,b_2,\ldots,b_\Lambda}$ be an increasing sequence of integers marking the repeated value intervals in $D_X$ such that $b_0 = 0$ and $x_{b_{i-1} + 1} = x_{b_{i-1} + 2} = \ldots = x_{b_i} = x'_i$.
By this definition, the allowable discretization positions are $d_i = \paren{x_{b_{i}} + x_{{b_i}+1}}/2$ for all \todo{$i = 1,2,\ldots,M-1$}.

@@

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm~\ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $\Lambda$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\State
\For {$v = 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u = 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $\Lambda$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

% \section{Experiments}
% \label{sec:experiments}

% This section describes experiments conducted to evaluate the proposed methods.
% All experiments were run on datasets from the publically available Irvine machine learning repository \citep{Lichman_2013}.
% Variables are labelled alphabetically in the order given on the dataset information webpage.
% Subscripts inducate the number of discretized intervals.
% Shaded nodes are originally discrete and unshaded nodes are originally continuous.

% For each dataset, the experiment consists of two parts.
% In the first part, the Bayesian network structure is known in advance, which is obtained by equal-width prediscretization and the K2 algorithm.
% In the second part, the Bayesian netowrk structure is not provided in advance, and it is learned during the discretization process.
